---
title: "Pacific calculus decontam"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(decontam)
library(data.table)
library(janitor)
library(tidyverse)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```


Load metadata
```{r load_metadata, eval = F}
#only need to run this once, then read in the file with the DNA concentration added
metadata <- fread("./05-Documentation.backup/oceana_metadata.tsv") %>%
  rename(Library_ID = LibraryID) %>%
  mutate(Library_ID = str_replace_all(Library_ID, ".SG1.1","")) %>%
  full_join(., fread("./05-Documentation.backup/oceana_meta_decontam.tsv") %>%
              select(-Lab) %>%
              rename(Library_ID = SampleID), by = "Library_ID") %>%
  mutate(Sample_or_Control = ifelse(SourceSink == "sink", "sample","control")) %>%
  mutate(Sample_or_Control = ifelse(Env == "blank", "control",Sample_or_Control)) %>%
  mutate(Sample_or_Control = ifelse(is.na(Sample_or_Control) & Type == "blank", "control",Sample_or_Control)) %>%
  mutate(Sample_or_Control = ifelse(is.na(Sample_or_Control) & Type == "bone", "control",Sample_or_Control)) %>%
  mutate(Sample_or_Control = ifelse(is.na(Sample_or_Control) & Type == "calculus","sample",Sample_or_Control)) %>%
  mutate(Type = ifelse(str_detect(Library_ID, "ARS"),"bone",Type))

# can't write into the Pacific folder, so write it here and move it to 05-Documentation.backup
# fwrite(metadata, "../strep_clades/00-documentation/oceana_metadata_merged.tsv", quote = F, sep = "\t", na = "NA")

```

```{r}
metadata <- fread("05-Documentation.backup/oceania_metadata_full.tsv") %>%
  # this sample has no information so we'll remove it
  filter(Library_ID != "HCLVMBCX2-3505-26-00-01_S26")

```


```{r poorly_preserved}
cfdp_fail <- fread("05-Documentation.backup/cfdp_fail_samples_20210413.txt") %>%
  rename(Library_ID = 1)

```

```{r}
# list all bones that are part of this study
# the ARS bones are a little different in the PCA, so we won't use them
bones <- metadata %>%
  filter(Type == "bone") %>%
  select(Library_ID, Lab, Type) %>%
  filter(!str_detect(Library_ID, "ARS|HPD"))

# list all samples and blanks from Otago
otago <- metadata %>%
  filter(Lab == "Otago") %>%
  select(Library_ID, Lab, Type) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13") %>%
  bind_rows(bones) %>%
  arrange(Library_ID) %>%
  pull(Library_ID)

# list all sampmles and blanks from Oklahoma
oklahoma <- metadata %>%
  filter(Lab == "Oklahoma") %>%
  select(Library_ID, Lab, Type) %>%
  filter(Type != "bone") %>%
  bind_rows(bones) %>%
  bind_rows(., metadata %>%
              filter(str_detect(Library_ID, "LIB")) %>% select(Library_ID, Lab, Type)) %>%
  pull(Library_ID)

# list all samples and blanks from Jena
jena <- metadata %>%
  filter(Lab == "Jena") %>%
  select(Library_ID, Lab, Type) %>%
  filter(Type != "bone") %>%
  bind_rows(bones) %>%
  pull(Library_ID)


```


```{r pca_plotting_functions}
# plotting PCA with colored dots
plot_pca <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(metadata_group, "_shapes", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("Library_ID") %>%
              inner_join(metadata %>%
                           select(Library_ID, Lab, Sample_or_Control,Type), by = "Library_ID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group, shape = metadata_group)) +
                        geom_point(size = 4) +
                        scale_colour_manual(values = metadata_group_colors) +
                        scale_shape_manual(values = metadata_group_shapes) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot)
}

# this function plots the PCA with the sample names instead of points
plot_pca_names <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("Library_ID") %>%
              inner_join(metadata %>%
                           select(Library_ID, Lab, Sample_or_Control,Type), by = "Library_ID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot_names <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group)) +
                        geom_text(aes(label = Library_ID), size = 3) +
                        scale_colour_manual(values = metadata_group_colors) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot_names)
}

```


## Genus

```{r load_genus_tables}
# read in the species table
rs_raw_gn <- fread("./05-Documentation.backup/oceania_malt_cRefSeq_summarized_genus_20201105.txt") %>%
  rename(Genus = 1) %>%
  full_join(., fread("./05-Documentation.backup/sources_MALT_cRefSeq_JFY_genus_summarized.txt") %>%
              rename(Genus = 1), by = "Genus") %>%
  full_join(., fread("./05-Documentation.backup/oceania_malt_blanks_cRefSeq_genus_summarized_20201130.txt") %>%
              rename(Genus = 1), by = "Genus") %>%
  full_join(., fread("./05-Documentation.backup/oceania_malt_bones_cRefSeq_genus_summarized_20201202.txt") %>%
              rename(Genus = 1), by = "Genus") %>%
  filter(Genus != "Homo")

# clean the file names
colnames(rs_raw_gn) <- gsub("MeganServer::", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".unmapped", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".SG1.1", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".SG1.2", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub("_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub("_S0_L000_R1_000.fastq.merged.prefixed.hg19unmapped", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub("_S0_L001_R1_001.fastq.truncated.prefixed.extractunmapped.bam", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub("_S0_L001_R1_000.fastq.merged.prefixed.hg1", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub("_S0_L000_R1_000.fastq.merged.prefixed.hg1", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".SG1.2_S0_L003_R1_001.fastq.combined.fq.prefixed.extrac.bam", "", colnames(rs_raw_gn))  
colnames(rs_raw_gn) <- gsub("_S0_L003_R1_001.fastq.combined.fq.prefixed.extrac.bam", "", colnames(rs_raw_gn)) 

# now do a pretend transpose to replace the NAs with 0s
rs_raw_gn <- rs_raw_gn %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = replace_na(Counts, 0)) %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts")

```


### Raw PCA
Before running decontam, let's plot the samples in a PCA to see if there are any that are clearly not good-quality
```{r raw_plot_nt}

genus_raw <- rs_raw_gn %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = Counts + 1) %>%
  pivot_wider(names_from = "Genus", values_from = "Counts") %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species_pca <- mixOmics::tune.pca(genus_raw, logratio = 'CLR')
tune.species_pca

# perform a PCA to see how the data cluster
genus_raw.pca <- mixOmics::pca(genus_raw, ncomp = 3, logratio = 'CLR')
plot(genus_raw.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

Type_colors <- c("black","blue","turquoise", "red")
Type_shapes <- c(19,17,15,9)

plot_pca(genus_raw.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca(genus_raw.pca, "PC1", "PC2", "Type")
plot_pca_names(genus_raw.pca,  "PC1", "PC2", "Sample_or_Control")

genus_raw_pass <- rs_raw_gn %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = Counts + 1) %>%
  pivot_wider(names_from = "Genus", values_from = "Counts") %>%
  anti_join(., cfdp_fail) %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species_pca <- mixOmics::tune.pca(genus_raw_pass, logratio = 'CLR')
tune.species_pca

# perform a PCA to see how the data cluster
genus_raw_pass.pca <- mixOmics::pca(genus_raw_pass, ncomp = 3, logratio = 'CLR')
plot(genus_raw_pass.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

plot_pca(genus_raw_pass.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca(genus_raw_pass.pca, "PC1", "PC2", "Type")


```

```{r outliers}
# HCLVMBCX2-3505-13-00-01_S13 appears to be a sample, although it's supposed to be a blank
# remove it from all further analyses

metadata_filt <- metadata %>%
  anti_join(., cfdp_fail) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13")

genus_raw_pass <- rs_raw_gn %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  pivot_wider(names_from = "Genus", values_from = "Counts") %>%
  anti_join(., cfdp_fail) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13")

```

### Otago

```{r otago_data_prep}

# select only Otago samples and controls
# note that all poorly-preserved samples are removed from genus_raw_pass although they're still in the otago list
otago_df <- genus_raw_pass %>%
  filter(Library_ID %in% otago) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# make a matrix for decontam
otago_matrix <- data.matrix(otago_df[2:ncol(otago_df)])
rownames(otago_matrix) <- otago_df$Library_ID

# make a metadata dataframe with the same samples 
otago_metadata <- otago_df %>%
  select(Library_ID) %>%
  left_join(., metadata, by = "Library_ID")

```

```{r otago_prevalence}
# Assign sample-type status to identify control samples
otago_metadata$is.control <- otago_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.otago_gn <- isContaminant(otago_matrix, method="prevalence", 
                                 neg=otago_metadata$is.control, 
                                 threshold = 0.5) # 0.69

# Check number of contaminants
table(contam.prev.otago_gn$contaminant)

# List of contaminants
contam.prev.otago_gn <- contam.prev.otago_gn %>%
  as_tibble() %>%
  mutate(Genus = row.names(contam.prev.otago_gn)) %>%
  select(Genus, everything())

contam.prev.otago_gn_list <- contam.prev.otago_gn %>%
  filter(contaminant == "TRUE") %>%
  arrange(Genus)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.otago_gn, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Otago genus')

```

```{r contaminants, eval = F}
contaminant_otago_gn <- contam.prev.otago_gn_list %>% select(Genus)
# fwrite(contaminant_otago_gn, file = "./05-Documentation.backup/contaminant_otago_gn.tsv", sep="\t", quote=F)
fwrite(contaminant_otago_gn, file = "../strep_clades/00-documentation/contaminant_otago_gn.tsv", sep="\t", quote=F)

```


### Oklahoma

```{r oklahoma_data_prep}
# select only Otago samples and controls
# note that all poorly-preserved samples are removed from genus_raw_pass although they're still in the otago list
oklahoma_df <- genus_raw_pass %>%
  filter(Library_ID %in% oklahoma) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# make a matrix for decontam
oklahoma_matrix <- data.matrix(oklahoma_df[2:ncol(oklahoma_df)])
rownames(oklahoma_matrix) <- oklahoma_df$Library_ID

# make a metadata dataframe with the same samples 
oklahoma_metadata <- oklahoma_df %>%
  select(Library_ID) %>%
  left_join(., metadata, by = "Library_ID")

```

```{r oklahoma_prevalence}
# Assign sample-type status to identify control samples
oklahoma_metadata$is.control <- oklahoma_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.oklahoma_gn <- isContaminant(oklahoma_matrix, method="prevalence", 
                                 neg=oklahoma_metadata$is.control, 
                                 threshold = 0.22)

# Check number of contaminants
table(contam.prev.oklahoma_gn$contaminant)

# List of contaminants
contam.prev.oklahoma_gn <- contam.prev.oklahoma_gn %>%
  as_tibble() %>%
  mutate(Genus = row.names(contam.prev.oklahoma_gn)) %>%
  select(Genus, everything())

contam.prev.oklahoma_gn_list <- contam.prev.oklahoma_gn %>%
  filter(contaminant == "TRUE") %>%
  arrange(Genus)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.oklahoma_gn, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')

```

```{r rad_contaminants}
contaminant_oklahoma_gn <- contam.prev.oklahoma_gn_list %>% select(Genus)
# fwrite(contaminant_oklahoma_gn, file = "./05-Documentation.backup/contaminant_oklahoma_gn.tsv", sep="\t", quote=F)
fwrite(contaminant_oklahoma_gn, file = "../strep_clades/00-documentation/contaminant_oklahoma_gn.tsv", sep="\t", quote=F)

```


```{r oklahoma_frequency, eval = F}
# Assign sample-type status to identify control samples
oklahoma_metadata$is.control <- oklahoma_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.freq.oklahoma_gn <- isContaminant(oklahoma_matrix, method="frequency", 
                                 conc=oklahoma_metadata$DNA_conc, 
                                 threshold = 0.99)

# Check number of contaminants
table(contam.freq.oklahoma_gn$contaminant)

# List of contaminants
contam.freq.oklahoma_gn <- contam.freq.oklahoma_gn %>%
  as_tibble() %>%
  mutate(Genus = row.names(contam.freq.oklahoma_gn)) %>%
  select(Genus, everything())

contam.freq.oklahoma_gn_list <- contam.freq.oklahoma_gn %>%
  filter(contaminant == "TRUE") %>%
  arrange(Genus)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.freq.oklahoma_gn, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Oklahoma genus')

```

```{r oklahoma_combined, eval = F}
# Assign sample-type status to identify control samples
oklahoma_metadata$is.control <- oklahoma_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.comb.oklahoma_gn <- isContaminant(oklahoma_matrix, method="combined", 
                                 conc=oklahoma_metadata$DNA_conc, 
                                 neg=oklahoma_metadata$is.control, 
                                 threshold = 0.9)

# Check number of contaminants
table(contam.comb.oklahoma_gn$contaminant)

# List of contaminants
contam.comb.oklahoma_gn <- contam.comb.oklahoma_gn %>%
  as_tibble() %>%
  mutate(Genus = row.names(contam.comb.oklahoma_gn)) %>%
  select(Genus, everything())

contam.comb.oklahoma_gn_list <- contam.comb.oklahoma_gn %>%
  filter(contaminant == "TRUE") %>%
  arrange(Genus)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.comb.oklahoma_gn, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Oklahoma genus')

```

### Jena

```{r jena_data_prep}
# select only Otago samples and controls
# note that all poorly-preserved samples are removed from genus_raw_pass although they're still in the otago list
jena_df <- genus_raw_pass %>%
  filter(Library_ID %in% jena) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# make a matrix for decontam
jena_matrix <- data.matrix(jena_df[2:ncol(jena_df)])
rownames(jena_matrix) <- jena_df$Library_ID

# make a metadata dataframe with the same samples 
jena_metadata <- jena_df %>%
  select(Library_ID) %>%
  left_join(., metadata, by = "Library_ID")

```

```{r jena_prevalence}
# Assign sample-type status to identify control samples
jena_metadata$is.control <- jena_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.jena_gn <- isContaminant(jena_matrix, method="prevalence", 
                                 neg=jena_metadata$is.control, 
                                 threshold = 0.022)

# Check number of contaminants
table(contam.prev.jena_gn$contaminant)

# List of contaminants
contam.prev.jena_gn <- contam.prev.jena_gn %>%
  as_tibble() %>%
  mutate(Genus = row.names(contam.prev.jena_gn)) %>%
  select(Genus, everything())

contam.prev.jena_gn_list <- contam.prev.jena_gn %>%
  filter(contaminant == "TRUE") %>%
  arrange(Genus)

# Look at the histogram - does the probability threshold need to be changed?
# ggplot(contam.prev.jena_gn, aes(x=p)) +
#   geom_histogram(bins=100) +
#   scale_y_log10() +
#   ggtitle('Prevalence - Species')

```

```{r jena_contaminants}
contaminant_jena_gn <- contam.prev.jena_gn_list %>% select(Genus)
# fwrite(contaminant_jena_gn, file = "./05-Documentation.backup/contaminant_jena_gn.tsv", sep="\t", quote=F)
fwrite(contaminant_jena_gn, file = "../strep_clades/00-documentation/contaminant_jena_gn.tsv", sep="\t", quote=F)

```

Combine all contaminant lists together into 1 and check for oral taxa
```{r}

pacific_gn_contaminants <- contaminant_otago_gn %>%
  mutate(otago = 1) %>%
  full_join(., contaminant_oklahoma_gn %>%
  mutate(okla = 1)) %>%
  full_join(., contaminant_jena_gn %>%
  mutate(jena = 1)) %>%
  arrange(Genus)

# fwrite(pacific_gn_contaminants, file = "./05-Documentation.backup/contaminant_pacific_ooj_gn.tsv", sep="\t", quote=F)
fwrite(pacific_gn_contaminants, file = "../strep_clades/00-documentation/contaminant_pacific_ooj_gn.tsv", sep="\t", quote=F)

```



## Species
```{r load_species_tables}
# read in the species table
rs_raw_sp <- fread("./05-Documentation.backup/oceania_malt_cRefSeq_summarized_species_20201105.txt") %>%
  rename(Species = 1) %>%
  full_join(., fread("./05-Documentation.backup/sources_MALT_cRefSeq_JFY_species_summarized.txt") %>%
              rename(Species = 1), by = "Species") %>%
  full_join(., fread("./05-Documentation.backup/oceania_malt_blanks_cRefSeq_species_summarized_20201130.txt") %>%
              rename(Species = 1), by = "Species") %>%
  full_join(., fread("./05-Documentation.backup/oceania_malt_bones_cRefSeq_species_summarized_20201202.txt") %>%
              rename(Species = 1), by = "Species") %>%
  filter(Species != "Homo sapiens")

# clean the file names
colnames(rs_raw_sp) <- gsub("MeganServer::", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".unmapped", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".SG1.1", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".SG1.2", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub("_L001_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub("_S0_L000_R1_000.fastq.merged.prefixed.hg19unmapped", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub("_S0_L001_R1_001.fastq.truncated.prefixed.extractunmapped.bam", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub("_S0_L001_R1_000.fastq.merged.prefixed.hg1", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub("_S0_L000_R1_000.fastq.merged.prefixed.hg1", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".SG1.2_S0_L003_R1_001.fastq.combined.fq.prefixed.extrac.bam", "", colnames(rs_raw_sp))  
colnames(rs_raw_sp) <- gsub("_S0_L003_R1_001.fastq.combined.fq.prefixed.extrac.bam", "", colnames(rs_raw_sp)) 

# now do a pretend transpose to replace the NAs with 0s
rs_raw_sp <- rs_raw_sp %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = replace_na(Counts, 0)) %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts")

```


### Raw PCA
Before running decontam, let's plot the samples in a PCA to see if there are any that are clearly not good-quality
```{r raw_plot_nt}

species_raw <- rs_raw_sp %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = Counts + 1) %>%
  pivot_wider(names_from = "Species", values_from = "Counts") %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species_pca <- mixOmics::tune.pca(species_raw, logratio = 'CLR')
tune.species_pca

# perform a PCA to see how the data cluster
species_raw.pca <- mixOmics::pca(species_raw, ncomp = 3, logratio = 'CLR')
plot(species_raw.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

Type_colors <- c("black","blue","turquoise", "red")
Type_shapes <- c(19,17,15,9)

plot_pca(species_raw.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca(species_raw.pca, "PC1", "PC2", "Type")
plot_pca_names(species_raw.pca,  "PC1", "PC2", "Sample_or_Control")

species_raw_pass <- rs_raw_sp %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = Counts + 1) %>%
  pivot_wider(names_from = "Species", values_from = "Counts") %>%
  anti_join(., cfdp_fail) %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.species_pca <- mixOmics::tune.pca(species_raw_pass, logratio = 'CLR')
tune.species_pca

# perform a PCA to see how the data cluster
species_raw_pass.pca <- mixOmics::pca(species_raw_pass, ncomp = 3, logratio = 'CLR')
plot(species_raw_pass.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

plot_pca(species_raw_pass.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca(species_raw_pass.pca, "PC1", "PC2", "Type")


```

```{r outliers_species}
# HCLVMBCX2-3505-13-00-01_S13 appears to be a sample, although it's supposed to be a blank
# remove it from all further analyses

species_raw_pass <- rs_raw_sp %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  pivot_wider(names_from = "Species", values_from = "Counts") %>%
  anti_join(., cfdp_fail) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13")

```

### Otago

```{r otago_data_prep}

# select only Otago samples and controls
# note that all poorly-preserved samples are removed from species_raw_pass although they're still in the otago list
otago_df <- species_raw_pass %>%
  filter(Library_ID %in% otago) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# make a matrix for decontam
otago_matrix <- data.matrix(otago_df[2:ncol(otago_df)])
rownames(otago_matrix) <- otago_df$Library_ID

# make a metadata dataframe with the same samples 
otago_metadata <- otago_df %>%
  select(Library_ID) %>%
  left_join(., metadata, by = "Library_ID")

```

```{r otago_prevalence}
# Assign sample-type status to identify control samples
otago_metadata$is.control <- otago_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.otago_sp <- isContaminant(otago_matrix, method="prevalence", 
                                 neg=otago_metadata$is.control, 
                                 threshold = 0.90) # 0.69

# Check number of contaminants
table(contam.prev.otago_sp$contaminant)

# List of contaminants
contam.prev.otago_sp <- contam.prev.otago_sp %>%
  as_tibble() %>%
  mutate(Species = row.names(contam.prev.otago_sp)) %>%
  select(Species, everything())

contam.prev.otago_sp_list <- contam.prev.otago_sp %>%
  filter(contaminant == "TRUE") %>%
  arrange(Species)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.otago_sp, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Otago species')

```

```{r contaminants}
contaminant_otago_sp <- contam.prev.otago_sp_list %>% select(Species)
# fwrite(contaminant_otago_sp, file = "./05-Documentation.backup/contaminant_otago_sp.tsv", sep="\t", quote=F)
fwrite(contaminant_otago_sp, file = "../strep_clades/00-documentation/contaminant_otago_sp.tsv", sep="\t", quote=F)

```


### Oklahoma

```{r oklahoma_data_prep}
# select only Otago samples and controls
# note that all poorly-preserved samples are removed from species_raw_pass although they're still in the oklahoma list
oklahoma_df <- species_raw_pass %>%
  filter(Library_ID %in% oklahoma) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# make a matrix for decontam
oklahoma_matrix <- data.matrix(oklahoma_df[2:ncol(oklahoma_df)])
rownames(oklahoma_matrix) <- oklahoma_df$Library_ID

# make a metadata dataframe with the same samples 
oklahoma_metadata <- oklahoma_df %>%
  select(Library_ID) %>%
  left_join(., metadata, by = "Library_ID")

```

```{r oklahoma_prevalence}
# Assign sample-type status to identify control samples
oklahoma_metadata$is.control <- oklahoma_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.oklahoma_sp <- isContaminant(oklahoma_matrix, method="prevalence", 
                                 neg=oklahoma_metadata$is.control, 
                                 threshold = 0.8)

# Check number of contaminants
table(contam.prev.oklahoma_sp$contaminant)

# List of contaminants
contam.prev.oklahoma_sp <- contam.prev.oklahoma_sp %>%
  as_tibble() %>%
  mutate(Species = row.names(contam.prev.oklahoma_sp)) %>%
  select(Species, everything())

contam.prev.oklahoma_sp_list <- contam.prev.oklahoma_sp %>%
  filter(contaminant == "TRUE") %>%
  arrange(Species)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.oklahoma_sp, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')

```

```{r oklahoma_contaminants}
contaminant_oklahoma_sp <- contam.prev.oklahoma_sp_list %>% select(Species)
# fwrite(contaminant_oklahoma_sp, file = "./05-Documentation.backup/contaminant_oklahoma_sp.tsv", sep="\t", quote=F)
fwrite(contaminant_oklahoma_sp, file = "../strep_clades/00-documentation/contaminant_oklahoma_sp.tsv", sep="\t", quote=F)

```


```{r oklahoma_frequency, eval = F}
# Assign sample-type status to identify control samples
oklahoma_metadata$is.control <- oklahoma_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.freq.oklahoma_sp <- isContaminant(oklahoma_matrix, method="frequency", 
                                 conc=oklahoma_metadata$DNA_conc, 
                                 threshold = 0.99)

# Check number of contaminants
table(contam.freq.oklahoma_sp$contaminant)

# List of contaminants
contam.freq.oklahoma_sp <- contam.freq.oklahoma_sp %>%
  as_tibble() %>%
  mutate(Species = row.names(contam.freq.oklahoma_sp)) %>%
  select(Species, everything())

contam.freq.oklahoma_sp_list <- contam.freq.oklahoma_sp %>%
  filter(contaminant == "TRUE") %>%
  arrange(Species)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.freq.oklahoma_sp, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Oklahoma species')

```

```{r oklahoma_combined, eval = F}
# Assign sample-type status to identify control samples
oklahoma_metadata$is.control <- oklahoma_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.comb.oklahoma_sp <- isContaminant(oklahoma_matrix, method="combined", 
                                 conc=oklahoma_metadata$DNA_conc, 
                                 neg=oklahoma_metadata$is.control, 
                                 threshold = 0.9)

# Check number of contaminants
table(contam.comb.oklahoma_sp$contaminant)

# List of contaminants
contam.comb.oklahoma_sp <- contam.comb.oklahoma_sp %>%
  as_tibble() %>%
  mutate(Species = row.names(contam.comb.oklahoma_sp)) %>%
  select(Species, everything())

contam.comb.oklahoma_sp_list <- contam.comb.oklahoma_sp %>%
  filter(contaminant == "TRUE") %>%
  arrange(Species)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.comb.oklahoma_sp, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Oklahoma species')

```

### Jena

```{r jena_data_prep}
# select only Otago samples and controls
# note that all poorly-preserved samples are removed from species_raw_pass although they're still in the jena list
jena_df <- species_raw_pass %>%
  filter(Library_ID %in% jena) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)

# make a matrix for decontam
jena_matrix <- data.matrix(jena_df[2:ncol(jena_df)])
rownames(jena_matrix) <- jena_df$Library_ID

# make a metadata dataframe with the same samples 
jena_metadata <- jena_df %>%
  select(Library_ID) %>%
  left_join(., metadata, by = "Library_ID")

```

```{r jena_prevalence}
# Assign sample-type status to identify control samples
jena_metadata$is.control <- jena_metadata$Sample_or_Control=="control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.jena_sp <- isContaminant(jena_matrix, method="prevalence", 
                                 neg=jena_metadata$is.control, 
                                 threshold = 0.57)

# Check number of contaminants
table(contam.prev.jena_sp$contaminant)

# List of contaminants
contam.prev.jena_sp <- contam.prev.jena_sp %>%
  as_tibble() %>%
  mutate(Species = row.names(contam.prev.jena_sp)) %>%
  select(Species, everything())

contam.prev.jena_sp_list <- contam.prev.jena_sp %>%
  filter(contaminant == "TRUE") %>%
  arrange(Species)

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.jena_sp, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')

```

```{r jena_contaminants}
contaminant_jena_sp <- contam.prev.jena_sp_list %>% select(Species)
# fwrite(contaminant_jena_sp, file = "./05-Documentation.backup/contaminant_jena_sp.tsv", sep="\t", quote=F)
fwrite(contaminant_jena_sp, file = "../strep_clades/00-documentation/contaminant_jena_sp.tsv", sep="\t", quote=F)

```

Combine all contaminant lists together into 1 and check for oral taxa
```{r}

pacific_sp_contaminants <- contaminant_otago_sp %>%
  mutate(otago = 1) %>%
  full_join(., contaminant_oklahoma_sp %>%
  mutate(okla = 1)) %>%
  full_join(., contaminant_jena_sp %>%
  mutate(jena = 1)) %>%
  arrange(Species)

# fwrite(pacific_sp_contaminants, file = "./05-Documentation.backup/contaminant_pacific_ooj_sp.tsv", sep="\t", quote=F)
fwrite(pacific_sp_contaminants, file = "../strep_clades/00-documentation/contaminant_pacific_ooj_sp.tsv", sep="\t", quote=F)

```












