---
title: "Pacific calculus clustering as in Quagliariello, et al. 2022 Nat Comm"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

## **NOTE** ## 
This can only be run on the cluster b/c of gfortran issues with getting SpiecEasi installed for NetCoMi.
Or not, nevermind b/c the optimal # of clusters is 1, and therefore no network analysis is necessary

```{r load_libraries, echo = F, message = F}
library(knitr)
library(data.table)
library(janitor)
library(rstatix)
library(vegan)
library(tidyverse)
library(patchwork)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

On the gap statistic: https://towardsdatascience.com/k-means-clustering-and-the-gap-statistics-4c5d414acd29

# contaminant tables 
```{r load_contam_lists}
# genus lists
cont_otago_gn <- fread("./05-Documentation.backup/contaminant_otago_gn.tsv")
cont_oklahoma_gn <- fread("./05-Documentation.backup/contaminant_oklahoma_gn.tsv")
cont_jena_gn <- fread("./05-Documentation.backup/contaminant_jena_gn.tsv")

cont_all_gn <- fread("./05-Documentation.backup/contaminant_pacific_ooj_gn.tsv")

# make sure the same contaminants are in the individual and combined tables (expect 874 - yes, same total)
cont_otago_gn %>%
  bind_rows(., cont_oklahoma_gn) %>%
  bind_rows(., cont_jena_gn) %>%
  distinct() %>%
  count()


# species lists
cont_otago_sp <- fread("./05-Documentation.backup/contaminant_otago_sp.tsv", sep = "\t")
cont_oklahoma_sp <- fread("./05-Documentation.backup/contaminant_oklahoma_sp.tsv", sep = "\t")
cont_jena_sp <- fread("./05-Documentation.backup/contaminant_jena_sp.tsv", sep = "\t")

cont_all_sp <- fread("./05-Documentation.backup/contaminant_pacific_ooj_sp.tsv", sep = "\t")

```

Load metadata
```{r load_metadata, eval = F}
#only need to run this once, then read in the file with the DNA concentration added
metadata <- fread("./05-Documentation.backup/oceana_metadata.tsv") %>%
  rename(Library_ID = Library_ID) %>%
  mutate(Library_ID = str_replace_all(Library_ID, ".SG1.1","")) %>%
  full_join(., fread("./05-Documentation.backup/oceana_meta_decontam.tsv") %>%
              select(-Lab) %>%
              rename(Library_ID = SampleID), by = "Library_ID") %>%
  mutate(Sample_or_Control = ifelse(SourceSink == "sink", "sample","control")) %>%
  mutate(Sample_or_Control = ifelse(Env == "blank", "control",Sample_or_Control)) %>%
  mutate(Sample_or_Control = ifelse(is.na(Sample_or_Control) & Type == "blank", "control",Sample_or_Control)) %>%
  mutate(Sample_or_Control = ifelse(is.na(Sample_or_Control) & Type == "bone", "control",Sample_or_Control)) %>%
  mutate(Sample_or_Control = ifelse(is.na(Sample_or_Control) & Type == "calculus","sample",Sample_or_Control)) %>%
  mutate(Type = ifelse(str_detect(Library_ID, "ARS"),"bone",Type))

# can't write into the Pacific folder, so write it here and move it to 05-Documentation.backup
# fwrite(metadata, "../strep_clades/00-documentation/oceana_metadata_merged.tsv", quote = F, sep = "\t", na = "NA")

```

```{r}
metadata <- fread("05-Documentation.backup/oceania_metadata_full.tsv") %>%
  # make the island names match the colors/shapes
  mutate(Island = str_replace_all(Island, "_"," "),
         Island = str_replace_all(Island, " Duff Islands",""),
         Island = str_replace_all(Island, "3000","3000 BP")) %>%
  # convert all numeric to character otherwise get an error about converting integers and characters
  mutate(DNA_conc = as.character(DNA_conc),
         Age_mean = as.character(Age_mean),
         Lat = as.character(Lat),
         Long = as.character(Long),
         seq_len_avg = as.character(seq_len_avg),
         gc_avg = as.character(gc_avg),
         No_reads_no_human = as.character(No_reads_no_human)) %>%
  mutate_all(na_if,"") %>%
  # now convert back to numeric
  mutate(DNA_conc = as.numeric(DNA_conc),
         Age_mean = as.numeric(Age_mean),
         Lat = as.numeric(Lat),
         Long = as.numeric(Long),
         seq_len_avg = as.numeric(seq_len_avg),
         gc_avg = as.numeric(gc_avg),
         No_reads_no_human = as.numeric(No_reads_no_human))

```


```{r poorly_preserved}
cfdp_fail <- fread("05-Documentation.backup/oceania_cuperdec_species_poor_samples.tsv") %>%
  rename(Library_ID = 1)

```

```{r}
# list all bones that are part of this study
# the ARS bones are a little different in the PCA, so we won't use them
bones <- metadata %>%
  filter(Type == "bone") %>%
  select(Library_ID, Lab, Type) %>%
  filter(!str_detect(Library_ID, "ARS|HPD"))

# list all samples and blanks from Otago
otago <- metadata %>%
  anti_join(., cfdp_fail) %>%
  filter(Lab == "Otago") %>%
  select(Library_ID, Lab, Type) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13") %>%
  arrange(Library_ID) %>%
  filter(Type != "blank") %>%
  pull(Library_ID)

# list all samples and blanks from Oklahoma
oklahoma <- metadata %>%
  anti_join(., cfdp_fail) %>%
  filter(Lab == "Oklahoma") %>%
  select(Library_ID, Lab, Type) %>%
  filter(Type != "bone") %>%
  filter(Type != "blank") %>%
  pull(Library_ID)

# list all samples and blanks from Jena
jena <- metadata %>%
  anti_join(., cfdp_fail) %>%
  filter(Lab == "Jena") %>%
  select(Library_ID, Lab, Type) %>%
  filter(Type != "bone") %>%
  filter(Type != "blank") %>%
  filter(!str_detect(Library_ID, "HPD")) %>%
  pull(Library_ID)

# list all samples to include
sample_list <- c(otago,oklahoma,jena)

```

```{r load_species_tables}
# read in the species table
rs_raw_sp <- fread("./05-Documentation.backup/oceania_malt_cRefSeq_summarized_species_20201105.txt") %>%
  rename(Species = 1) %>%
  filter(Species != "Homo")

# clean the file names
colnames(rs_raw_sp) <- gsub("MeganServer::", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".unmapped", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".SG1.1", "", colnames(rs_raw_sp)) 
colnames(rs_raw_sp) <- gsub(".SG1.2", "", colnames(rs_raw_sp)) 

# now do a pretend transpose to replace the NAs with 0s
rs_raw_sp <- rs_raw_sp %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = replace_na(Counts, 0)) %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts")

```


```{r outliers}
# HCLVMBCX2-3505-13-00-01_S13 appears to be a sample, although it's supposed to be a blank
# remove it from all further analyses

metadata_filt <- metadata %>%
  anti_join(., cfdp_fail) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13")

species_raw_pass <- rs_raw_sp %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  pivot_wider(names_from = "Species", values_from = "Counts") %>%
  anti_join(., cfdp_fail) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13") %>%
  # remove all taxa that have to entries anymore
  pivot_longer(!Library_ID, names_to = "Species", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  adorn_totals(where = "col") %>%
  select(Total, everything()) %>%
  filter(Total != 0) %>%
  select(-Total) %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  pivot_wider(names_from = "Species", values_from = "Counts")

```


### Decontam all together

```{r all_table}
# note that all poorly-preserved samples are removed from species_raw_pass although they're still in the lab lists

# select all samples and remove all contaminants
# save row names
rownames_sp_all_filt <- species_raw_pass %>%
  filter(Library_ID %in% sample_list) %>%
  pivot_longer(!Library_ID, names_to = "Species", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  anti_join(., cont_all_sp %>%
              select(Species), by = "Species") %>%
  adorn_totals(where = "col") %>%
  filter(Total != 0) %>%
  select(-Total) %>%
  pull(Species)

sp_all_filt <- species_raw_pass %>%
  filter(Library_ID %in% sample_list) %>%
  pivot_longer(!Library_ID, names_to = "Species", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  anti_join(., cont_all_sp %>%
              select(Species), by = "Species") %>%
  adorn_totals(where = "col") %>%
  filter(Total != 0) %>%
  select(-Total) %>%
  # remove the totals attribute so can make column sums later if necessary by converting to matrix and back
  as.matrix(., rownames = "Species") %>%
  as_tibble(.) %>%
  mutate(Species = rownames_sp_all_filt) %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = as.numeric(Counts)) %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts")

# remove the totals attribute so can make column sums later if necessary
# attr(sp_all_filt, "totals") <- NULL

```


Remove species that are present at <0.005% relative abundance in each lab
```{r each_plot_abund005_filt}

# species present at <0.005% per lab
sp_abund005_each <- sp_all_filt %>%
  # Jena
  select(matches(c("Species",jena))) %>%
  adorn_totals(where = "col") %>%
  mutate(Percent = Total / sum(Total) * 100) %>%
  filter(Percent <= 0.005) %>%
  select(Species) %>%
  full_join(., sp_all_filt %>%
            # Oklahoma
            select(matches(c("Species",oklahoma))) %>%
            adorn_totals(where = "col") %>%
            mutate(Percent = Total / sum(Total) * 100) %>%
            filter(Percent <= 0.005) %>%
            select(Species)) %>%
  full_join(., sp_all_filt %>%
            # Otago
            select(matches(c("Species",otago))) %>%
            adorn_totals(where = "col") %>%
            mutate(Percent = Total / sum(Total) * 100) %>%
            filter(Percent <= 0.005) %>%
            select(Species))


sp_each_abund005_filt_1 <- sp_all_filt %>%
  anti_join(., sp_abund005_each) %>%
  pivot_longer(!Species, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = Counts + 1) %>%
  pivot_wider(names_from = "Species", values_from = "Counts") %>%
  column_to_rownames("Library_ID")

```

```{r}
# Run a pca to get a clr-transformed table for input to clustering
sp_each_abund005_filt_1.pca <- mixOmics::pca(sp_each_abund005_filt_1)

# The clr-transformed matrix from the mixOmics PCA above
sp_each_abund005_filt_1_clr <- sp_each_abund005_filt_1.pca$X %>%
  as.data.frame(.)

sp_each_abund005_filt_1_clr_t <- sp_each_abund005_filt_1_clr %>%
  rownames_to_column("Library_ID") %>%
  pivot_longer(!Library_ID, names_to = "Species", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  column_to_rownames("Species")

```

```{r}

# cluster samples with the Gap statsitic, like in Quagliariello

gp_cluster_out <- cluster::clusGap(sp_each_abund005_filt_1_clr_t, d.power = 2, FUN = kmeans, K.max = 10, B = 500) # nstart = 20, B = 500 for real run, spaceH0 = "scaledPCA"

gp_cluster_out$Tab %>%
  as.data.frame.matrix() %>%
  mutate(No_clusters = c(1:10)) %>%
  select(No_clusters, everything())

sp_gap_plot <- gp_cluster_out$Tab %>%
  as.data.frame.matrix() %>%
  mutate(No_clusters = c(1:10)) %>%
  select(No_clusters, everything()) %>%
  ggplot(., aes(x = No_clusters, y = gap)) +
         geom_line() +
         geom_point() +
         geom_errorbar(aes(ymin = gap-SE.sim, ymax = gap+SE.sim), width = 0.2) +
         theme_minimal(base_size = 12) +
         theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12)) +
         theme(axis.text.y = element_text(size = 10)) +
         theme(panel.grid.minor = element_blank()) +
         scale_x_continuous(breaks=scales::pretty_breaks(n=10)) +
         scale_y_continuous(breaks=scales::pretty_breaks(n=10)) +
        ylab("Gap statistic") +
        xlab("Number of clusters") +
        ggtitle("Species table")

sp_gap_plot

```

## Genus
```{r load_genus_tables}
# read in the species table
rs_raw_gn <- fread("./05-Documentation.backup/oceania_malt_cRefSeq_summarized_genus_20201105.txt") %>%
  rename(Genus = 1) %>%
  filter(Genus != "Homo")

# clean the file names
colnames(rs_raw_gn) <- gsub("MeganServer::", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".unmapped", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".SG1.1", "", colnames(rs_raw_gn)) 
colnames(rs_raw_gn) <- gsub(".SG1.2", "", colnames(rs_raw_gn)) 

# now do a pretend transpose to replace the NAs with 0s
rs_raw_gn <- rs_raw_gn %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  mutate(Counts = replace_na(Counts, 0)) %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts")

# make sure there are no genera with no counts in any sample
rs_raw_gn %>%
  adorn_totals(where = "col") %>%
  select(Genus, Total) %>%
  arrange(Total) %>%
  slice_min(Total)

```

```{r all_table}
genus_raw_pass <- rs_raw_gn %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  pivot_wider(names_from = "Genus", values_from = "Counts") %>%
  anti_join(., cfdp_fail %>%
              as_data_frame(.) %>%
              rename(Library_ID = 1)) %>%
  filter(Library_ID != "HCLVMBCX2-3505-13-00-01_S13") %>%
  # remove all taxa that have to entries anymore
  pivot_longer(!Library_ID, names_to = "Genus", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  adorn_totals(where = "col") %>%
  filter(Total != 0) %>%
  select(-Total) %>%
  pivot_longer(!Genus, names_to = "Library_ID", values_to = "Counts") %>%
  pivot_wider(names_from = "Genus", values_from = "Counts")


# note that all poorly-preserved samples are removed from genus_raw_pass although they're still in the lab lists

# select all samples and remove all contaminants, and remove all genera with no counts in any sample
gn_all_filt <- genus_raw_pass %>%
  filter(Library_ID %in% sample_list) %>%
  pivot_longer(!Library_ID, names_to = "Genus", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  anti_join(., cont_all_gn %>%
              select(Genus), by = "Genus") %>%
  adorn_totals(where = "col") %>%
  filter(Total != 0) %>%
  select(-Total)


```

```{r}
# Run a pca to get a clr-transformed table for input to clustering
gn_all_filt.pca <- mixOmics::pca(gn_all_filt)

# The clr-transformed matrix from the mixOmics PCA above
gn_all_filt_clr <- gn_all_filt.pca$X %>%
  as.data.frame(.)

gn_all_filt_clr_t <- gn_all_filt_clr %>%
  rownames_to_column("Library_ID") %>%
  pivot_longer(!Library_ID, names_to = "Species", values_to = "Counts") %>%
  pivot_wider(names_from = "Library_ID", values_from = "Counts") %>%
  column_to_rownames("Species")

```

```{r}

# cluster samples with the Gap statsitic, like in Quagliariello
# set.seed(501) # 4, 250, 501
gp_cluster_out <- cluster::clusGap(gn_all_filt_clr_t, d.power = 2, FUN = kmeans, K.max = 10, B = 500) # nstart = 20, B = 500 for real run, spaceH0 = "scaledPCA"

gp_cluster_out$Tab %>%
  as.data.frame.matrix() %>%
  mutate(No_clusters = c(1:10)) %>%
  select(No_clusters, everything())

gn_gap_plot <- gp_cluster_out$Tab %>%
  as.data.frame.matrix() %>%
  mutate(No_clusters = c(1:10)) %>%
  select(No_clusters, everything()) %>%
  ggplot(., aes(x = No_clusters, y = gap)) +
         geom_line() +
         geom_point() +
         geom_errorbar(aes(ymin = gap-SE.sim, ymax = gap+SE.sim), width = 0.2) +
         theme_minimal(base_size = 12) +
         theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12)) +
         theme(axis.text.y = element_text(size = 10)) +
         theme(panel.grid.minor = element_blank()) +
         scale_x_continuous(breaks=scales::pretty_breaks(n=10)) +
         scale_y_continuous(breaks=scales::pretty_breaks(n=10)) +
        ylab("Gap statistic") +
        xlab("Number of clusters") +
        ggtitle("Genus table")

gn_gap_plot

```

```{r}

gap_plot <- sp_gap_plot / gn_gap_plot +
  plot_layout()

gap_plot

# ggsave("./06-publication/supplemental_figures/Sup_fig_2/S2_pacific_gap_statistic.pdf", plot = gap_plot, device = "pdf",
#        scale = 1, width = 8, height = 7, units = c("in"), dpi = 300)
# 
# ggsave("./06-publication/supplemental_figures/Sup_fig_2/S2_pacific_gap_statistic.png", plot = gap_plot, device = "png",
#        scale = 1, width = 8, height = 7, units = c("in"), dpi = 300)

```


